Where Do Events Come From?

### 2018-08-06 paper.tex
...

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop:
from a sensor input source and back to an actuator output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{
    interrupt service routines,
    reactive programming,
    sleep modes,
    standby,
    synchronous/asynchronous execution
}

@@

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
%As illustrated in Figure~\ref{fig.env},
Peripherals can be encapsulated as a
single component, \emph{the environment}, which connects with the application
in an event loop:
the application sits idle until the environment awakes it on the occurrence of
an input;
the application reacts to the input and generates back one or more outputs
which affect the environment;
the application becomes idle and the loop restarts.

@@

The environment is typically implemented in a host language (e.g., C) and
controls the main event loop, invoking entry points into the reactive language
runtime on the occurrence of inputs, and also receiving output calls from it,
both through a documented API.
%
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript as commands and listening for values as
subscriptions~\cite{frp.elm.ports}.

The event-based interface between the application and the environment is
arguably inevitable and also happens at the appropriate layer, since it
connects the application with the operating system resources through a
non-invasive API.
%
However, this separation reveals the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
\CEU targets resource-constrained architectures, such as Arduino-compatible
microcontrollers, in which applications run in the bare metal, without
operating system support.
In this context, device drivers are typically libraries compiled together with
the main program.

The lack of an OS opens an opportunity to explore a tighter integration between
the application and its device drivers at the language level, resulting in an
overall simplicity and tractability of the system.
%
In particular, \CEU is already amenable to a simple static analysis to detect
race conditions which we extended in this work to also encompass ISRs.
Applications can share data buffers with drivers to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU also provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts and turn off peripherals completely.
For instance, we can delimit the scope of a driver similarly to a
lexically-scoped variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, we can put the microcontroller to
sleep automatically at optimal sleeping modes after each input reaction.

We implemented an extensible runtime that exposes hooks for the interrupts and
power manager, which we validated in two microcontrollers: an
\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We also implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
The applications built on top of these drivers show optimal energy savings due
to automatic standby.

\begin{comment}
Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
wireless devices, which also provides asynchronous events with a compile-time
data race detector~\cite{wsn.nesc}, and automatic energy management triggered
from idle CPU states~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive programming model of \CEU,
and refined them with stronger guarantees due to lexical scope and automatic
finalization.
\end{comment}

@@

\CEU~\cite{ceu.sensys13} is a Esterel-based~\cite{ceu.tecs17} reactive
programming language targeting resource-constrained embedded systems, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms such as \code{spawn} and \code{await} (to create and suspend
    lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

\begin{comment}
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
\end{comment}

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
%The \code{par/or} is regarded as an \emph{orthogonal preemption
%primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
%in order to affect each other.

@@

Drivers

Our initial tests show optimal energy savings for applications in idle states.

>>>>

\begin{abstract}
In reactive and event-based systems, execution is guided by an external
environment that generates inputs to the application and consumes outputs from
it.
Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the programming experience in this domain.
%
Nevertheless, the environment is typically prefabricated in a host language and
the very central concept of events is implemented externally to the reactive
language.
%
In this work, we propose an interrupt handler primitive for a reactive language
targeting embedded systems in order to take control of the whole event loop,
from input to output.
%
We propose the new asynchronous primitive in the context of the synchronous
language \CEU and discuss how they synergize to avoid race conditions at
compile time, support lexically-scoped drivers, and provide automatic standby
for applications.
%and hot swapping of device drivers
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{
    interrupt service routines,
    reactive programming,
    standby,
    synchronous/asynchronous execution
}

@@

Reactive applications interact continuously and in real time with the external
world through hardware peripherals such as sensors and actuators (e.g.,
buttons, displays, timers, etc.).
%
These interactions are typically represented in software as input events
flowing from the peripherals to the application and as output events flowing
from the application to the peripherals.
%
%As illustrated in Figure~\ref{fig.env},
Peripherals can be encapsulated as a
single component, \emph{the environment}, which connects with the application
through an event loop:
the application sits idle until the environment awakes it on the occurrence of
an input;
the application reacts to the input and generates back outputs
which affect the environment;
the application becomes idle and the loop restarts.

@@

Reactive languages provide dedicated syntax and semantics to deal with events
and greatly simplify the development of applications in this domain.
%
However, the environment is typically implemented in a host language (e.g., C)
and controls the main event loop, invoking entry points into the reactive
language runtime on the occurrence of inputs, and also receiving output calls
from it.
%, both through a documented API.
%
As examples, Esterel~\cite{esterel.ieee91} relies on C for passing events
between the environment and the running program~\cite{esterel.book.compiling},
while Elm~\cite{frp.elm} uses the concept of \emph{ports}, which allows sending
out values to JavaScript as commands and listening for values as
subscriptions~\cite{frp.elm.ports}.

The event-based interface between the reactive language and the environment is
arguably inevitable and also happens at the appropriate layer, since it
connects the application with the operating system (OS) resources through a
non-invasive API.
%
However, this separation reveals the environment as a rigid system component
that evolves in separate from the application.
%
It also requires two languages and the programmer has to deal with multiple
syntaxes, incompatible type systems, and different address spaces.
%
Furthermore, in the context of embedded systems, a proper host OS may even be
absent or lacking enough device drivers, which requires more low-level
intervention from the application.

In this work, we propose interrupt service routines (ISRs) as an asynchronous
primitive for the synchronous reactive language \CEU~\cite{ceu.sensys13}.
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
\CEU targets resource-constrained architectures, such as Arduino-compatible
microcontrollers, in which applications run in the bare metal, without
OS support.
%In this context, device drivers are typically libraries compiled together with
%the main program.
%
%The lack of an OS opens opportunities to explore a tighter integration between
%the application and its device drivers at the language level, resulting in an
%overall simplicity and tractability of the system.
%
%In particular, \CEU is already amenable to a simple static analysis to detect
%race conditions which we extended in this work to also encompass ISRs.
Applications can share data buffers with drivers to avoid unnecessary copying,
with some static guarantee that no data races will occur.
%
\CEU provides a lexical finalization mechanism that we adopt in drivers to
properly disable interrupts and turn off peripherals completely.
%For instance, we can delimit the scope of drivers similarly to lexically-scoped
%variables.
%
Finally, the synchronous semantics of \CEU enforces that applications react
to inputs in bounded time and remain in idle states susceptible to standby.
With the help of drivers and a power manager, we can put the microcontroller to
sleep automatically at optimal sleeping modes after each input reaction.
%
%We implemented an extensible runtime that exposes hooks for the ISRs and
%power manager, which we validated in two microcontrollers: an
%\emph{8-bit AVR/ATmega} and a \emph{32-bit ARM/Cortex-M0}.
%
We implemented drivers for a variety of peripherals, such as GPIO, A/D
converter, USART, SPI, and nRF24L01 transceiver.
%
The applications built on top of these drivers show optimal energy savings due
to automatic standby.

Our work is largely inspired by TinyOS~\cite{wsn.tos}, a low-power OS for
wireless devices, which also provides asynchronous events with compile-time
data race detection~\cite{wsn.nesc}, and automatic energy management
in idle CPU states~\cite{wsn.icem}.
%
We adapted these ideas to the structured reactive programming model of \CEU,
and refined them with stronger guarantees due to its support for lexically
scoped lines of execution with automatic finalization.

@@

\CEU~\cite{ceu.sensys13} is a Esterel-based
%~\cite{ceu.tecs17}
reactive
programming language targeting resource-constrained embedded systems, with the
characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use lexically-scoped structured control
    mechanisms such as \code{spawn} and \code{await} (to create and suspend
    lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}

Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating}. %,rp.rescala,sync_async.cooperative

@@

The \code{par/or}, which stands for \emph{parallel-or}, is a lexical
composition that terminates as soon as one of the trails terminates, but which
also automatically aborts and finalizes the other trail(s).
%
In the example, when the button is pressed (ln~5), not only the toggling loop
will be aborted (ln~10--15), but the \code{finalize} clause will turn the LED
off unconditionally (ln~8), since its enclosing block goes out of scope.
%
The \code{par/or} is regarded as an \emph{orthogonal preemption
primitive}~\cite{esterel.preemption} because the trails need not to be tweaked
in order to affect each other.

@@

We extend the synchronous language \CEU with asynchronous interrupt service
routines (ISRs) to take control of the whole event loop in reactive systems,
from input to output.
%
ISRs empower reactive applications to also implement their own device drivers
and self-generate inputs, bypassing the need for a host environment.
%
However, asynchronous execution confronts the well-behaved semantics of
synchronous languages with possible race conditions.
%
In order to mitigate this threat, we extend the static concurrency checks of
\CEU to detect data races between ISRs and the application.

Our approach for writing drivers take advantage of the lexical structured
mechanisms of \CEU, such as parallel compositions and abortion of lines of
execution, to offer stronger guarantees.
%
For instance, the visibility of a driver can be delimited to a scoped block
with the guarantee that its resources will not leak after usage.
%
The same mechanism permits that drivers signal the power manager that its
peripheral is no longer required, which allows applications to use optimal
sleep modes to save energy automatically.
%
Our initial tests show optimal energy savings for applications in idle states.
